---
title: "IPUMSR workshop notebook"
output: html_notebook
---

1. Install the package

```{r}
install.packages("ipumsr")
```

2. Load the library

```{r}
library(ipumsr)
```

3. Set API key on environment

```{r}
set_ipums_api_key("59cba10d8a5da536fc06b59d10166c045d174963a84c0d0a1cd77825", save = TRUE,)
```

4. Exploring the NHGIS datasets

```{r}
nhgis_data <- get_metadata_nhgis("datasets")
nhgis_data
```

5. Filtering the results

```{r}
install.packages("dplyr")
```

```{r}
library(dplyr)
nhgis_data |>
filter(group == "2023 American Community Survey") |>
select(name, description)
```

6. Exploring the dataset metadata

```{r}
ds_meta <- get_metadata_nhgis(dataset = "2019_2023_ACS5a")
ds_meta
```

7. Exploring all tables

```{r}
ds_meta$data_tables
```

8. Finding the correct tables

```{r}
install.packages("stringr")
```

```{r}
library(stringr)
ds_meta$data_tables |>
filter(str_detect(description, "Tenure"))
```

9. Define a data extract

```{r}
my_extract <- define_extract_nhgis(
  description = "Homeownership by race, PA",
  datasets = ds_spec(
    "2019_2023_ACS5a",
    data_tables = c("B25003B", "B25003H", "B25003I"),
    geog_levels = "county"
    ),
  shapefiles = "us_county_2023_tl2023", # explore available shapefiles in IPUMS website
)
```

10. Submiting and downloading

```{r}
my_extract <- submit_extract(my_extract)
my_extract <- wait_for_extract(my_extract)
my_extract <- download_extract(my_extract)
```

11. Reading the files

```{r}
nhgis_data <- read_nhgis("nhgis0013_csv.zip")
spatial_data <- read_ipums_sf("nhgis0013_shape.zip")
```

12. Exploring our data

```{r}
colnames(nhgis_data)
```

```{r}
ipums_var_info(nhgis_data$ASTBE001)
```

13. Calculating the proportion of owners by race

```{r}
nhgis_data <- nhgis_data |>
mutate(
ho_prop_white = ASTHE002 / ASTHE001,
ho_prop_black = ASTBE002 / ASTBE001,
ho_prop_hispanic = ASTIE002 / ASTIE001
)|>
mutate(
ho_ratio_white_black = ho_prop_white / ho_prop_black,
ho_ratio_white_hispanic = ho_prop_white / ho_prop_hispanic,
)|>
select(YEAR, STATEA, STATE, COUNTYA, COUNTY, GISJOIN, starts_with("ho_ratio")) |>
  filter(STATEA==42)
```

14. Joining the data

```{r}
for_mapping <- ipums_shape_inner_join(nhgis_data, spatial_data, by = "GISJOIN")
```

15. Preparing mapping packages

- ggplot allows to create multiple types of visualizations
```{r}
install.packages("ggplot2")
library(ggplot2)
```

- `sf` is a package to manage spatial formats
```{r}
install.packages("sf")
library(sf)
```

16. Creating a simple map

```{r}
ggplot(data = for_mapping, aes(fill = ho_ratio_white_hispanic)) + 
  geom_sf()
```

17. Improving the map

```{r}
ggplot(data = for_mapping, aes(fill = ho_ratio_white_hispanic)) + 
  geom_sf() + 
  scale_fill_distiller(palette = "YlOrRd", 
                       direction = 1) + 
  labs(title = "Homeownership ratio white/hispanic, PA, 2023",
       caption = "Data source: 2022 5-year ACS, US Census Bureau",
       fill = "Ratio") + 
  theme_void()
```

17. Creating an interactive map with `mapgl()`

```{r}
install.packages("mapgl")
library(mapgl)
maplibre()
```

```{r}
pa_map <- maplibre(bounds=for_mapping)
```

18. Mapping

```{r}
pa_map |> 
  add_fill_layer(
  id = "homeowner",
  source = for_mapping,
  fill_color = interpolate(
    column = "ho_ratio_white_hispanic",
    values = c(1, 3),
    stops = c("lightyellow", "darkorange"),
    na_color = "lightgrey"
  ),
  fill_opacity = 0.7
 ) |> 
  add_legend(
    "Homeownership ratio white/hispanic, PA, 2023",
    values = c(1, 3),
    colors = c("lightyellow", "darkorange")
  )
```

